\chapter{Attribution rules}\label{attr}

\section{Regular attribution rules}\label{reg-attr}
An attribution rule consists of a tree expression and a block. When
attribution rules are executed, the abstract syntax tree is traversed
depth-first. For each visited node of the abstract syntax tree, all
attribution rules are executed whose tree expressions match the visited
node.

Attributes rules are executed in pre- or postorder in respect to the
traversal of the subnodes. By default, rules are executed in preorder.
The keywords \keyword{pre} and \keyword{post} can be used to specify the
order.

When multiple rules match, they are ordered first by arity (in preorder
those rules with varying arity are executed first, in postorder rules
with varying arity come last) and second by the order of appearance in
the source.

\begin{grammar}
   \nonterminal{attribution-rules}
      \produces \lexkeyword{attribution} \lexkeyword{rules}
	 \lextoken{\{} \nonterminal{attributions}
	 \lextoken{\}} \\
   \nonterminal{attributions}
      \produces \nonterminal{attribution} \\
      \produces \nonterminal{attributions} \nonterminal{attribution} \\
   \nonterminal{attribution}
      \produces \nonterminal{conditional-tree-expressions}
	 \lextoken{->} \nonterminal{block} \\
      \produces \nonterminal{conditional-tree-expressions}
	 \lextoken{->} \lexkeyword{pre} \nonterminal{block} \\
      \produces \nonterminal{conditional-tree-expressions}
	 \lextoken{->} \lexkeyword{post} \nonterminal{block} \\
\end{grammar}

\noindent
Example: In the following example we have three attribution rules
that focus on operator nodes with the operator
\lstinline!"compound_statement"!. The first rule is executed
first. Afterwards, the second rule is executed whenever a
compound statement is nested within another compound statement.
Then all the attribution rules are executed that match the subnodes.
Finally, the third rule is executed last as it is a postfix rule.

\begin{lstlisting}
   ("compound_statement" *) as block -> {
      block.vars = {};
      block.level = 0;
      block.up = null;
   }
   ("compound_statement" *) as inner_block
         in ("compound_statement" *) as outer_block -> {
      inner_block.level = outer_block.level + 1;
      inner_block.up = outer_block;
   }
   ("compound_statement" *) as block in ("translation_unit" *) as root -> post {
      foreach (varname, entry) in (block.vars) {
      if (entry.used) {
	 if (entry.minblock != block) {
	    println(location(entry.decl) & ": variable " & varname &
	       " should be moved into " & location(entry.minblock));
	 }
      } else {
	 println(location(entry.decl) & ": unused variable " & varname);
      }
   }
\end{lstlisting}

\section{Named attribution rules}\label{named-atrules}

Named attribution rules are similar to regular attribution rules but
they are not implicitly executed (see \ref{xorder}). The given name
is bound to a function which executes the associated attribute rule
set. The function takes one optional parameter that specifies the
root node of the to be traversed tree. Otherwise, if no parameter is
given, \ident{root} is taken (see \ref{predefined}).
Within this rule set, \ident{root}\index{root} is bound to the
to be attributed abstract syntax tree.

\begin{grammar}
   \nonterminal{attribution-rules}
      \produces \lexkeyword{attribution} \lexkeyword{rules}
	 \nonterminal{identifier}
	 \lextoken{\{} \nonterminal{attributions}
	 \lextoken{\}}
\end{grammar}

\endinput
