\chapter{Lexical elements}

\section{Source files}

Astl program source files\index{source file} must end in the
suffix ``.ast''\index{suffix} which is to be omitted when
source files are refered to.

\section{Character set}\index{character set}

Astl program sources are sequences of octets that are interpreted as
ASCII characters. Octets within string literals, program text literals,
and comments are not interpreted except to find the end of the lexical
construct or to find embedded references within program text literals.

Case is significant.

\section{Lines and columns}

To provide locations\index{location} for error messages,
lines\index{line} and columns\index{column} are tracked while reading a
program source\index{source file}. The newline\index{newline} character
(ASCII 10) is interpreted as line terminator\index{line terminator}
and considered otherwise as being equivalent to the space\index{space}
character (ASCII 32). Tabs\index{tab} (ASCII 9) are similarly treated as
space characters but cause the current column to be advanced to the next
multiply of 8 plus 1.

\section{Comments}\index{comment}

Comments can be delimited by ``/*'' and ``*/''. Nested comments
are not supported. Alternatively, comments start with ``//'' and
are ended by the next line terminator\index{line terminator}.
Comments are handled like space characters.

\section{Tokens}\index{token}

Each program source is converted into a sequence of tokens during
the lexical analysis\index{lexical analysis}. Tokens can be
delimiters\index{delimiter} consisting of one or more special
characters\index{special character}, reserved keywords\index{keyword},
identifiers\index{identifier}, or literals\index{literal}. Tokens end if
the next character is a space or if the next character cannot be added
to it.

Example: The character sequence \lstinline!a[i2]+=exists f{i2}? 12: existsf!
consists of the following tokens:

\begin{tabular}{l l}
   \lstinline!a! & an identifier \\
   \lstinline![! & a single-character delimiter \\
   \lstinline!i2! & an identifier \\
   \lstinline!]! & a single-character delimiter \\
   \lstinline!+=! & a compound delimiter \\
   \lstinline!exists! & a keyword \\
   \lstinline!f! & an identifier \\
   \lstinline!{! & a single-character delimiter \\
   \lstinline!i2! & an identifier \\
   \lstinline!}! & a single-character delimiter \\
   \lstinline!?! & a single-character delimiter \\
   \lstinline!12! & a decimal literal \\
   \lstinline!:! & a single-character delimiter \\
   \lstinline!existsf! & an identifier \\
\end{tabular}

\section{Single-character delimiters}\index{delimiter}

Following delimiters consist of one character only and cannot
be part of a longer token which is not a literal:

\bigskip
\noindent
\begin{tabular}{l l l l}
   \lstinline!(! & \lstinline!)! & \lstinline!{! & \lstinline!}! \\
   \lstinline![! & \lstinline!]! & \lstinline!^! & \lstinline!*! \\
   \lstinline!.! & \lstinline!,! & \lstinline!?! & \lstinline!:! \\
   \lstinline!;! \\
\end{tabular}

\section{Compound delimiters}\index{delimiter}

Following delimiter consists of one or multiple characters:

\bigskip
\noindent
\begin{tabular}{l l l l}
   \lstinline!-! & \lstinline!->! & \lstinline!--! & \lstinline!-=! \\
   \lstinline!+! & \lstinline!+=! & \lstinline!++! \\
   \lstinline!&! & \lstinline!&&! & \lstinline!&=! \\
   \lstinline!<! & \lstinline!<=! \\
   \lstinline!>! & \lstinline!>=! \\
   \lstinline/!/ & \lstinline/!=/ \\
   \lstinline!=! & \lstinline!=~! & \lstinline!==! \\
   \lstinline!||! \\
\end{tabular}

\section{Keywords}\index{keyword}

Following keywords are reserved. They cannot be used as identifiers.

\bigskip
\noindent
\input{keywords}

\section{Identifiers}\index{identifier}

Identifiers begin with a letter, i.e. ``A'' to ``Z'' and ``a'' to ``z'',
or an underscore ``\_'', and are optionally continued with a sequence
of more letters, underscores, or decimal digits ``0'' to ``9''.

\section{Literals}\index{literal}

Literals can be decimal literals, string literals, regular expressions,
and program text literals.

\subsection{Decimal literals}\index{decimal literal}\index{literal!decimal}

Decimal literals begin with a decimal digit\index{digit} ``0'' to ``9'' and
optionally more digits. Decimal constants are unsigned and can
be of arbitrary size.

\subsection{String literals}\index{string literal}\index{literal!string}

String literals are delimited by ``"''. Backslashes\index{backslash},
i.e. ``\lstinline!\!'',
are escape characters, i.e. they remove the special meaning of
the following character or allow to insert special characters into a
string:

\bigskip

\noindent
\begin{tabular}{| l l l |}
\hline
new-line & NL (LF) & \lstinline!\n! \\
horizontal tab & HT & \lstinline!\t! \\
vertical tab & VT & \lstinline!\v! \\
backspace & BS & \lstinline!\b! \\
form feed & FF & \lstinline!\f! \\
alert & BEL & \lstinline!\a! \\
backslash & \backslashSY{} & \lstinline!\\! \\
question mark & ? & \lstinline!\?! \\
single quote & ' & \lstinline!\'! \\
double quote & " & \lstinline!\"! \\
octal number & \textit{ooo} & \lstinline!\!\textit{ooo} \\
\hline
\end{tabular}

\bigskip

\noindent
(This covers the set of \nonterminal{simple-escape-sequence} and
\nonterminal{octal-escape-sequence} of the C language as defined
by the ISO standard 9899-1999.)

Examples: \lstinline!"Hello, world"! represents the text ``Hello, world'',
\lstinline!"\""! represents the double quote ``"'',
\lstinline!"\\"! represents the backslash ``\lstinline!\!'',
and \lstinline!"Two\nlines"! represents ``Two'', followed by
a line terminator, and ``lines''.

\subsection{Regular expressions}\label{regexlit}
\index{regular expression}\index{literal!regular expression}

A regular expression is a literal that is enclosed by the
delimiters ``\lstinline!m{!'' and ``\lstinline!}!''. Regular
expressions are interpreted by the \textit{Perl Compatible
Regular Expression} library (PCRE).\footnote{See
\url{http://www.pcre.org} and
\url{http://www.pcre.org/current/doc/html/pcre2syntax.html}. The
options \textit{PCRE2\_ALT\_BSUX}, \textit{PCRE2\_UCP},
and \textit{PCRE2\_UTF} are set.} Regular expressions
must be encoded in UTF-8.

Example: \lstinline!m{[a-zA-Z_][a-zA-Z_0-9]*}!

\subsection{Program text literals}
\index{literal!program text}\index{program text literal}

A program text literal begins with ``\lstinline!q{!'' and
is ended by a balanced ``\lstinline!}!''.\footnote{This
type of literal is perhaps familiar to those knowing Perl.
However, its semantics follows that of the ``\lstinline!qq\{...\}!''
construct in Perl which allows interpolation.}
These literals are designed to support the generation of program text within
the print rules.\index{print rule} They may include embedded
references to variables or embedded expressions that are
interpolated when the literal is evaluated. As in
string literals, the backslash ``\lstinline!\!'' may be used to
escape following characters. Braces may be used without preceding
backslashes within a program text literal as long as they are
balanced.

Examples: \lstinline!q{{}}! is equivalent to \lstinline!q{\{\}}!.

\subsubsection{Embedded variable references}

An embedded variable reference begins with the character
``\lstinline!$!''%stopzone
and is followed by an identifier. Spaces between
``\lstinline!$!''%stopzone
and the identifier following it are not permitted.

\subsubsection{Embedded expressions}

An embedded expression begins with the character sequence
``\lstinline!${!'', %stopzone%stopzone
is followed by an arbitrary expression, and closed by
``\lstinline!}!''.

\subsubsection{Interpretation of multi-line literals}

Program text literals may extend over multiple lines. In this
case, initial and trailing white space are to some extent ignored
and the indentations of the individual lines are interpreted
relatively to each other.

Example: Following program text literal generates an
if-else-statement construct that references the variables
\ident{condition}, \ident{then\_statement}, and \ident{else\_statement}.
The white space between the opening ``\lstinline!{!''
and ``if'' is ignored. Similarly, the trailing whitespace between
\lstinline!$else_statement! %stopzone
and the closing
``\lstinline!}!''
is skipped.
However, the relative indentations are noted, i.e.
``\lstinline!$then_statement!'' %stopzone
is to be indented by three additional spaces.

\begin{lstlisting}
   q{
      if ($condition)
         $then_statement
      else
         $else_statement
   };
\end{lstlisting}

\endinput
